---
title: "squdpy_nes_obtain_data"
author: "Daniel Dimitrov"
date: "10/8/2021"
output: html_document
---

This .Rmd is used to convert the processed anndata object from Squidpy to
appropriately formatted and normalized Seurat objects.
Further, I run the neighbourhood enrichment analysis from Squidpy, which seeks
to identify co-localizing cells using a graph structure.

## Load Required Libs
```{r setup, include=FALSE}
require(tidyverse)
require(reticulate)
require(Seurat)
require(magrittr)
require(liana)
```

```{python pysetup, include=FALSE}
import scanpy as sc
import pandas as pd
import numpy as np
import squidpy as sq
import anndata2ri
print(sq.__version__) # should be >1.1
```


## Obtain data and enrichment
```{python get_adata_nes}
"""""
Title: Function to return the EM and do NES with Squidpy

params:
    adata - anndata
    coord_type - type of coordinate system
    cluster_key - variable where clustering is stored
    spatial_key - where spatial coordinates are stored - for some reason relevant for neighbourhood analysis
    kwargs - passed to plot
    
    
returns: list with processed adata and Neighbourhood Enrichment Scores (NES)
"""""
def get_adata_nes(adata, cluster_key, spatial_key = "spatial", **kwargs):
    sq.gr.spatial_neighbors(adata, coord_type="generic", spatial_key=spatial_key)
    sq.gr.nhood_enrichment(adata, cluster_key=cluster_key)
    sq.pl.nhood_enrichment(adata, cluster_key=cluster_key, **kwargs)
    
    # nes
    cats = adata.obs[cluster_key].cat.categories
    nes_key = "{}_nhood_enrichment".format(cluster_key)
    nes = pd.DataFrame(adata.uns[nes_key]['zscore'], index=cats, columns=cats)
    
    both = dict(adata = adata, nes = nes)
    return(both)
```

### Function to Convert to Seurat
```{r adata_to_seurat, warning = FALSE}
#' Function to Convert adata to Seurat object
#' @param adata anndata object as obtain from the function above
#' @returns a Seurat object
adata_to_seurat <- function(adata){
    exprs <- t(as.matrix(py$adata$X))
    colnames(exprs) <- py$adata$obs_names$to_list()
    rownames(exprs) <- py$adata$var_names$to_list()
    # Create the Seurat object
    seurat_object <- CreateSeuratObject(exprs)
    # Set the expression assay
    seurat_object <- SetAssayData(seurat_object, "counts", exprs)
    # Add observation metadata
    seurat_object <- AddMetaData(seurat_object, py$adata$obs)
    return(seurat_object)
}
```

## Convert and Save Datasets
```{python seq_squidpy_res}
# seqFISH
# https://squidpy.readthedocs.io/en/stable/auto_tutorials/tutorial_seqfish.html#sphx-glr-auto-tutorials-tutorial-seqfish-py
adata = sq.datasets.seqfish()
adata = adata[adata.obs["celltype_mapped_refined"]!='Low quality']
seq_adata_nes = get_adata_nes(adata, cluster_key="celltype_mapped_refined", method="ward")
```


### seqFISH Format and Write Object
```{r seq_format_save, warning=FALSE}
# Get Seq Seurat
seq_seurat <- adata_to_seurat(py$seq_adata_nes$adata)

# format meta and remove low quality cells
seq_seurat@meta.data %<>% 
    filter(celltype_mapped_refined!="Low quality") %>%
    mutate(celltype_mapped_refined = gsub("[/]", ".", celltype_mapped_refined)) %>%
    mutate(seurat_clusters = 
               as.factor(str_replace_all(celltype_mapped_refined,
                                         " ", "\\."))) 
seq_seurat <- subset(seq_seurat, cells=rownames(seq_seurat@meta.data))
Idents(seq_seurat) <- seq_seurat@meta.data$seurat_clusters
levels(Idents(seq_seurat))

# run normalization etc
seq_seurat %<>%
    Seurat::SCTransform(verbose = FALSE) %>%
    Seurat::FindVariableFeatures() %>%
    Seurat::RunPCA(., verbose = FALSE) %>%
    Seurat::RunUMAP(., dims = 1:30, verbose = FALSE)

seq_seurat
saveRDS(seq_seurat, "data/input/spatial/fishes/seqFISH_seurat.RDS")

# get and save UNFORMATTED NES results (clusters to be rename and filtered)
nes <- py$seq_adata_nes$nes %>%
        py_to_r()
saveRDS(nes, "data/input/spatial/fishes/seqFISH_nes.RDS")
```


### merFISH Format and Write Object
```{python mer_squidpy_res}
# merFISH
# https://squidpy.readthedocs.io/en/stable/auto_tutorials/tutorial_merfish.html#sphx-glr-auto-tutorials-tutorial-merfish-py
adata = sq.datasets.merfish()
adata = adata[adata.obs["Cell_class"]!='Ambiguous']
mer_adata_nes = get_adata_nes(adata, cluster_key="Cell_class", spatial_key="spatial3d", method="single", cmap="inferno", vmin=-50, vmax=100)
```

```{r mer_format_save, warning=FALSE}
# Get MER Seurat
mer_seurat <- adata_to_seurat(py$mer_adata_nes$adata)

# format meta and remove low ambigious
mer_seurat@meta.data %<>%
    filter(Cell_class!="Ambiguous") %>%
    mutate(seurat_clusters = 
               str_replace_all(Cell_class, " ", "\\.") %>%
               as.factor())
mer_seurat <- subset(mer_seurat, cells=rownames(mer_seurat@meta.data))
Idents(mer_seurat) <- mer_seurat@meta.data$seurat_clusters
levels(Idents(mer_seurat))

# run normalization etc
mer_seurat %<>% 
    Seurat::SCTransform(verbose = FALSE) %>%
    Seurat::FindVariableFeatures() %>%
    Seurat::RunPCA(., verbose = FALSE) %>%
    Seurat::RunUMAP(., dims = 1:30, verbose = FALSE) 

# save object
mer_seurat
saveRDS(mer_seurat, "data/input/spatial/fishes/merFISH_seurat.RDS")

# get and save UNFORMATTED NES results (clusters to be rename and filtered)
nes <- py$mer_adata_nes$nes %>%
        py_to_r()

saveRDS(nes, "data/input/spatial/fishes/merFISH_nes.RDS")
```


